---
title: "Civil War Battles and Railroad Junction Proximities"
subtitle: "DSAN 6750 / PPOL 6805: GIS for Spatial Data Science"
authors:
  - name: Will Corbin
    email: wzc3@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
bibliography: GIS_Manuscript.bib
title-block-banner: "#E69F0095"
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: true
prefer-html: true
---

## Introduction

Several previous studies have found robust relationships between spatial properties of a country's **capital city** and that country's propensity for **conflict** and **misgovernance**.

Perceptions of this linkage also have an effect on "coup-proofing" decisions made by national governments. A recent BBC interview with Equatorial Guinea's President Teodoro Obiang, for example, highlighted this as a factor behind his decision to relocate the capital city:

> It's the remoteness of Oyala that makes it so appealing to President Obiang. In a rare interview he described how rebels had recently plotted a seaborne assault on his palace in the current capital, Malabo. 'We need a secure place for my government and for future governments. That's why we have created Oyala, to guarantee the government of Equatorial Guinea.' [@sackur_equatorial_2012]

This case is far from exceptional, as an even more recent *Washington Post* article points out with respect to Myanmar's decision to move its capital from Yangon to Naypyidaw:

> Analysts have described the decision as motivated by a desire to secure the military’s seat of power from any threat of protests or invasions. [@berger_myanmars_2021]

Most of these studies, however, are based on observations of **conflict events**. In this study, we study the more fundamental variable of a capital's distance from the **population centroid** of the country.

## Literature Review

## Methodology

## Exploratory Data Analysis (EDA)

```{r}
#| label: imports
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(units) |> suppressPackageStartupMessages()
cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
#| label: load junction data
junction_df <- read_csv("Junction.csv", show_col_types = FALSE)

junction_df |> head()
```

```{r}
#| label: map junction data
junction_sf <- junction_df |> sf::st_as_sf(
  coords=c("X","Y"),
  crs=4326
) |> sf::st_transform(3857)
mapview(junction_sf)

```

```{r}
#| label: load battle data
battle_df <- read_csv("nps_battles.csv", show_col_types = FALSE)
battle_df <- battle_df %>% 
  mutate(
    significance_numeric = case_when(
      significance == 'A' ~ 4,
      significance == 'B' ~ 3,
      significance == 'C' ~ 2,
      significance == 'D' ~ 1
    )
  )

battle_df |> head()

```

```{r}
battle_df$significance_numeric

```

```{r}
#| label: map battle data
battle_sf <- battle_df |> sf::st_as_sf(
  coords=c("long","lat"),
  crs=4326
) |> sf::st_transform(3857)
mapview(battle_sf,zcol="result", cex="significance_numeric", label="result")

```


We then construct an **area-normalized** measure of capital-centroid distance $\text{dist}^{\textsf{AN}}$, using the formula

$$
\text{dist}^{\textsf{AN}}_i = \text{dist}_i / \sqrt{\text{area}_i}.
$$

A plot of this measure by country looks as follows:

```{r}
#| label: plot-area
merged_area_sf <- readRDS("merged_area_sf.rds")
mapview(merged_area_sf, zcol="scaled_dist")
```

## Hypothesis Testing (Regression)


## Discussion

## Conclusion

