---
title: "Civil War Battles and Railroad Junction Proximities"
subtitle: "DSAN 6750 / PPOL 6805: GIS for Spatial Data Science"
authors:
  - name: Will Corbin
    email: wzc3@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
title-block-banner: "#E69F0095"
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: true
prefer-html: true
---

## Introduction

The US Civil War was one of the first wars that attempted to leverage the use of railroads. Controlling railroad locations and having easy access contributed to the Union's ability to win the war, especially becuase most battles 

## Literature Review

## Methodology

## Exploratory Data Analysis (EDA)

```{r}
#| label: imports
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(units) |> suppressPackageStartupMessages()
cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
#| label: load junction data
junction_df <- read_csv("Junction.csv", show_col_types = FALSE)

junction_df |> head()
```

```{r}
#| label: map junction data
junction_sf <- junction_df |> sf::st_as_sf(
  coords=c("X","Y"),
  crs=4326
) |> sf::st_transform(3857)
mapview(junction_sf)

```

```{r}
#| label: load battle data
battle_df <- read_csv("nps_battles.csv", show_col_types = FALSE)
battle_df <- battle_df %>% 
  mutate(
    significance_numeric = case_when(
      significance == 'A' ~ 4,
      significance == 'B' ~ 3,
      significance == 'C' ~ 2,
      significance == 'D' ~ 1
    )
  )

battle_df |> head()

```


```{r}
#| label: map battle data
battle_sf <- battle_df |> sf::st_as_sf(
  coords=c("long","lat"),
  crs=4326
) |> sf::st_transform(3857)
mapview(battle_sf,zcol="result", cex="significance_numeric", label="result")

```



## Hypothesis Testing (Regression)


## Discussion

## Conclusion

