---
title: "Civil War Battles and Railroad Junction Proximities"
subtitle: "DSAN 6750 / PPOL 6805: GIS for Spatial Data Science"
authors:
  - name: Will Corbin
    email: wzc3@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
title-block-banner: "#964B00"
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: true
prefer-html: true
---

## Introduction

The US Civil War was one of the first wars that attempted to leverage the use of railroads. Controlling railroad locations and having easy access contributed to the Union's ability to win the war, especially because most battles took place in the US South. Railroads are able to be utilized to transport troops and supplies, and by doing so provide crucial support for an army that is not on its home soil.

The question I would like to explore:
Is the Union more likely to win battles that are located in close proximity to railroad junctions?

Null hypothesis:
The likelihood that an Union victory occurs at the given point s_i is directly proportional to the density of railroad junctions at that point.

Alternative hypothesis:
There exists a population density threshold p* such that Union victories occur at locations s_i* with railroad junction density greater than p* more often than what would be expected solely on the basis of this railroad junction density.

## Roadmap

- Determine where the railroad junctions are located (specifically in the US South). 
- Determine where Civil War battles took place.
- Analyze distances from battle sites to railroad junctions.
- Compare intensity functions of railroad junctions and battle sites.
- Run an experiment that randomizes some element of battle sites location or railroad junction locations.
- Examine possible spatial effect of railroad junctions on the results of battles, specifically battles that resulted in an Union victory.
- Is there anything we can test regarding the significance of the battle?



## Literature Review

## Methodology

## Exploratory Data Analysis (EDA)

```{r}
#| label: imports
library(tidyverse) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(mapview) |> suppressPackageStartupMessages()
library(units) |> suppressPackageStartupMessages()
cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
#| label: load junction data
junction_df <- read_csv("Junction.csv", show_col_types = FALSE)

junction_df |> head()
```

```{r}
#| label: map junction data
junction_sf <- junction_df |> sf::st_as_sf(
  coords=c("X","Y"),
  crs=4326
) |> sf::st_transform(3857)
mapview(junction_sf)

```

```{r}
#| label: load battle data
battle_df <- read_csv("nps_battles.csv", show_col_types = FALSE)
battle_df <- battle_df %>% 
  mutate(
    significance_numeric = case_when(
      significance == 'A' ~ 4,
      significance == 'B' ~ 3,
      significance == 'C' ~ 2,
      significance == 'D' ~ 1
    )
  )

battle_df |> head()

```


```{r}
#| label: map battle data
battle_sf <- battle_df |> sf::st_as_sf(
  coords=c("long","lat"),
  crs=4326
) |> sf::st_transform(3857)
mapview(battle_sf,zcol="result", cex="significance_numeric", label="result")

```

## Hypothesis Testing




## Discussion

## Conclusion

